### 2017-05-30 doc/categories.tex

Figure~\ref{fig:omission-imaginet-quotient} shows a similar analysis
contrasting {\sc LM} with the {\sc Textual} pathway of {\sc
  Imaginet}. The first observation is that the range of values of the
log ratios is narrow, indicating that the differences between these
two networks regarding which grammatical categories they are sensitive
to is less pronounced than when comparing {\sc Visual} to {\sc
  Textual}. While the size of the effect is weak, there also seems to
be a tendency for the {\sc Textual} model to pay relatively more
attention to lexical and less to structural dependencies, compared to
{\sc LM}: it may be that the {\sc Visual} pathway pulls {\sc Textual}
in this direction by sharing word embeddings with it.

@@

Models that utilize the sequential structure of natural-language 
have the capacity to interpret the same word-type differently depending on
the context. The omission score distributions in Section \ref{sec:omitimaginet} 
show that in the case of {\sc Imaginet} the 
pathways are differentially sensitive to content vs.\ function
words. This may be either just due to purely lexical features or the model 
may actually learn to pay more attention to the same word type in appropriate
contexts. This section investigates to what extent {\sc Visual} and {\sc Textual}
discriminate between occurrences of a given word in different positions and 
grammatical functions. The analysis we described here takes \label{edit:beyonlexicalgeneral}
the omission scores as input data, therefore it can be potentially applied 
to any architecture for which the omission scores can be computed. However,
the presented analysis and results regarding word positions can only be meaningful
for Recurrent Neural Networks as they compute their representations sequentially and are not
limited by fixed window sizes.\footnote{CNNs with multi-word filters
and tree-structured recursive neural networks do not incrementaly build representations
of sentences in a left-to-right or right-to-left fashion. 
Bi-directional RNNs, however, are affected by word-order and can potentially
learn to handle the same word in different positions differently. \label{edit:foot}}

We fit four L2-penalized linear regression models which predict the omission 
scores per token with the following predictor variables:

\begin{enumerate}
	\item {\sc LR word}: word type
	\item {\sc LR +dep}: word type, dependency label and their interaction 
	\item {\sc LR +pos}: word type, position (binned as {\sc first, second, third, middle,
	antepenult, penult, last}) and their interaction
	\item {\sc LR full}: word type, dependency label, position, word:dependency interaction, 
	word:position interaction
\end{enumerate}

\noindent We split the 5-thousand-image portion of MSCOCO validation data into two parts, fit the 
models (regularized via L2 penalty) on the first part, and compute the 
proportion of variance explained on the second part. 
For comparison we also train a version of {\sc Imaginet},
where the GRU in the {\sc Visual} pathway is replaced by a simple vector-summation operation
over word-embeddings and thus does not have access to sequential cues. We refer to this
model as {\sc Sum}.

@@

Table~\ref{tab:lr-r2} shows the proportion of variance $R^2$ in omission
scores explained by the linear regression with the different predictors.
Figure~\ref{fig:rsquared} offers a different view of the data, showing
the increase or decrease in $R^2$ for the models per pathway relative to {\sc LR
  +pos}.  The results reveal that adding additional information on
top of lexical features in case of {\sc Sum} does increase the
explained variance slightly, which is probably due to the unseen words
in the held out set. More interestingly, there is a sizeable
increase in $R^2$ between {\sc LR +pos} and {\sc LR full} in
case of {\sc Visual}, suggesting that 
the omission scores in case of {\sc Visual} depend on the words'
grammatical function in sentences, even after controling for word
identity and linear position.

The raw $R^2$ scores show that for the {\sc Textual} and {\sc LM}
models the word-type predicts the omission-score to much smaller
degree compared to {\sc Visual}, and the increase in $R^2$ shows that
the position of the word adds considerable amount of information for
these language models: this is not surprising considering that the omission
scores are measured with respect to the final activation state. 
Less predictably, for the {\sc Visual} model,
dependency labels are more informative than linear position,
hinting at the importance of syntactic structure beyond linear order.

@@

\begin{figure}
\centering
  \includegraphics[scale=0.35]{position-new.pdf}
\caption{Proportion of variance in omission scores explained by the
  linear regression models
 for {\sc Sum}, {\sc Visual} and {\sc Textual} pathways, relative to
 regressing on word identity and position only. }
\label{fig:rsquared}
\end{figure}

@@

In order to find out some of the specific syntactic configurations leading to
increase in $R^2$ between {\sc LR word} and {\sc LR dep} predictors
in case of {\sc Visual}, we next considered all word types with
occurrence counts of at least 100 and ranked them according to how much
better, on average, {\sc LR +dep} predicted their omission scores 
compared to {\sc LM word}.

Figure~\ref{fig:top_words} shows the per-dependency 
omission-score distributions for the five top ranked words plus the word {\it  water}. 
There are clear and large differences in how these words
impact the network's representation depending on what grammatical
function they fulfill. They all have large omission scores when they
occur as {\sc nsubj} (nominal subject) or {\sc root}, likely due to the fact that these
grammatical functions typically have a large contribution to the
complete meaning of a sentence.  Conversely, all have small omission 
scores when appearing as {\sc conj} (conjunct): this is probably because in this position
they share their contribution with the first, often more important,
member of the conjunction (e.g. {\it A cow and its baby eating
  grass}).

The pattern for {\sc nn} (nominal modifier) is a bit more complicated: for
four of the words shown (as well as for most other words not shown in
the figure), the score is very low in this grammatical function--presumably 
because most words contribute less to the sentence meaning when used  
as modifiers than as heads (e.g.\ {\it a clock tower}). However,
for the words {\it zebra} and {\it water}, omission scores are high
when they act as a nominal modifier {\sc nn}. This appears due to two reasons:
\begin{enumerate}
\item For {\it zebra}, there are frequent erroneous parses such as {\it
    zebra/{\sc nn} browsing/{\sc root}} instead of {\it zebra/{\sc nsubj} browsing/{\sc
    root}}. The network does not make this mistake, and treats these
    occurrences of {\it zebra} according to its importance as {\sc nsubj}.
\item {\it Water} as a modifier often changes the meaning of its head in
 a visually salient way: e.g.\ {\it water fall, water balloon, water scene, water skiing}, and
  thus the network learns that this particular word is important in the modifier position.
\end{enumerate}

\subsubsection{Sensitivity to information structure}
\label{subsec:information-struct}

@@

As observed in Section \ref{sec:beyondlexical}, 
adding extra information about the position of words 
explains more of the variance in case of {\sc Visual} and especially {\sc Textual}.
Figure \ref{fig:posrqs} shows the coefficients corresponding to the
position variables in {\sc LM full}. Since the omission scores 
are measured at the end-of-sentence token, the expectation is that 
for {\sc Textual}, as a language model,  
the words appearing closer to the end of the sentence would have a
stronger effect on the omission scores. This seems to be confirmed by
the plot as the coefficients up until the \emph{penultimate} are all negative.

For the {\sc Visual} model it is less clear what to
expect: on the one hand due to their chain structure,  
RNNs are better at keeping track of
short-distance rather than long-distance dependencies and thus we can expect 
tokens in positions closer to the end of the sentence to be more important. 
On the other hand in English the information structure (or pragmatic structure) 
of a sentence is expressed via linear ordering: the {\sc topic} of a 
sentence appears sentence-initially, and the {\sc comment} follows. 
In the context of other text types such as dialog or multi-sentence \label{edit:topiccomment}
narrative structure, we would expect {\sc comment} to often be more
important than {\sc topic} as {\sc comment} will often 
contain new information in these cases. In our setting of image captions 
however, it is the {\sc topic} that typically contains the most important 
objects depicted in the image, e.g. {\it {\underline{two zebras}} are grazing in tall grass on a savannah.}
Thus, for the task of predicting features of the visual scene, it would 
be advantageous to detect the topic of the sentence and up-weight its
importance in the final meaning representation. Figure \ref{fig:posrqs}
appears to support this hypothesis and the network does 
learn to pay more attention to  words appearing
sentence-initially. This effect seems to be to some extent mixed with the recency
bias of RNNs as perhaps indicated by the relatively high coefficient of the {\it last}
position for {\sc Visual}.

>>>>

Figure~\ref{fig:omission-imaginet-quotient} shows a similar analysis
contrasting {\sc LM} with the {\sc Textual} pathway of {\sc
  Imaginet}. The first observation is that the range of values of the
log ratios is narrow, indicating that the differences between these
two networks regarding which grammatical categories they are sensitive
to is less pronounced than when comparing {\sc Visual} to {\sc
  Textual}. While the size of the effect is weak, there also seems to
be a tendency for the {\sc Textual} model to pay relatively more
attention to content and less to function words, compared to
{\sc LM}: it may be that the {\sc Visual} pathway pulls {\sc Textual}
in this direction by sharing word embeddings with it.

Most of our findings up to this point conform reasonably well to prior
expectations about effects that  particular learning objectives should
have. This fact serves to validate our methods. In the next section we
go on to investigate less straightforward patterns.

@@

Models that utilize the sequential structure of language 
have the capacity to interpret the same word-type differently depending on
the context. The omission score distributions in Section \ref{sec:omitimaginet} 
show that in the case of {\sc Imaginet} the 
pathways are differentially sensitive to content vs.\ function
words. In principle, this may be either just due to purely lexical features or the model 
may actually learn to pay more attention to the same word type in appropriate
contexts. This section investigates to what extent {\sc Visual} and {\sc Textual}
discriminate between occurrences of a given word in different positions and 
grammatical functions. The analysis we described here takes \label{edit:beyonlexicalgeneral}
the omission scores as input data, therefore it can be potentially applied 
to any architecture for which the omission scores can be computed. However,
the presented analysis and results regarding word positions can only be meaningful
for Recurrent Neural Networks as they compute their representations sequentially and are not
limited by fixed window sizes.\footnote{CNNs with multi-word filters
and tree-structured recursive neural networks do not incrementaly build representations
of sentences in a left-to-right or right-to-left fashion. 
Bi-directional RNNs, are sensitive by word-order and can potentially
learn to handle the same word in different positions differently. \label{edit:foot}}

We fit four L2-penalized linear regression models which predict the omission 
scores per token with the following predictor variables: 
\begin{enumerate}
	\item {\sc LR word}: word type
	\item {\sc LR +dep}: word type, dependency label and their interaction 
	\item {\sc LR +pos}: word type, position (binned as {\sc first, second, third, middle,
	antepenult, penult, last}) and their interaction
	\item {\sc LR full}: word type, dependency label, position, word:dependency interaction, 
	word:position interaction
\end{enumerate}

\noindent We split the 5-thousand-image portion of MSCOCO validation data into two parts, fit the 
models on the first part, and compute the 
proportion of variance explained on the second part. 
For comparison we also use the {\sc Sum} model which composes word
embeddings via summation, and uses the same loss function as {\sc
  Visual}. This model is unable to encode information about word
order, and thus is a good baseline here as we investigate the
sensitivity of the networks to positional and structural cues.

@@

Table~\ref{tab:lr-r2} shows the proportion of variance $R^2$ in omission
scores explained by the linear regression with the different predictors.
Figure~\ref{fig:rsquared} offers a different view of the data, showing
the increase or decrease in $R^2$ for the models relative to {\sc
  LR~+pos}.  
The results reveal that adding additional information on
top of lexical features in case of {\sc Sum} does increase the
explained variance slightly, which is most likely due to the unseen words
in the held out set. More interestingly, there is a sizeable
increase in $R^2$ between {\sc LR~+pos} and {\sc LR~full} in
case of {\sc Visual}, suggesting that 
the omission scores in case of {\sc Visual} depend on the words'
grammatical function in sentences, {\it even after controling for word
identity and linear position.}

The raw $R^2$ scores show that for the {\sc Textual} and {\sc LM}
models the word-type predicts the omission-score to much smaller
degree compared to {\sc Visual}, and the increase in $R^2$ shows that
the position of the word adds considerable amount of information for
these language models: this is not surprising considering that the omission
scores are measured with respect to the final activation state, and
given the fact that in a language model the recent history is most
important for accurate prediction. Less expectedly, for the {\sc Visual} model,
dependency labels are more informative than linear position,
hinting at the importance of syntactic structure beyond linear order.

@@

\begin{figure}
\centering
  \includegraphics[scale=0.35]{position-new.pdf}
\caption{Proportion of variance in omission scores explained by the
  linear regression models
 for {\sc Sum}, {\sc LM}, {\sc Visual} and {\sc Textual}, relative to
 regressing on word identity and position only. }
\label{fig:rsquared}
\end{figure}

@@

In order to find out some of the specific syntactic configurations leading to
increase in $R^2$ between the {\sc LR~word} and {\sc LR~dep} predictors
in case of {\sc Visual}, we next considered all word types with
occurrence counts of at least 100 and ranked them according to how much
better, on average, {\sc LR~+dep} predicted their omission scores 
compared to {\sc LM~word}.

Figure~\ref{fig:top_words} shows the per-dependency 
omission-score distributions for seven top ranked words.
There are clear and large differences in how these words
impact the network's representation depending on what grammatical
function they fulfill. They all have large omission scores when they
occur as {\sc nsubj} (nominal subject) or {\sc root}, likely due to the fact that these
grammatical functions typically have a large contribution to the
complete meaning of a sentence.  Conversely, all have small omission 
scores when appearing as {\sc conj} (conjunct): this is probably because in this position
they share their contribution with the first, often more important,
member of the conjunction, for example: {\it A cow and its baby eating
  grass}. 
% The pattern for {\sc nn} (nominal modifier) is a bit more complicated: for
% four of the words shown (as well as for most other words not shown in
% the figure), the score is very low in this grammatical function--presumably 
% because most words contribute less to the sentence meaning when used  
% as modifiers than as heads (e.g.\ {\it a clock tower}). However,
% for the words {\it zebra} and {\it water}, omission scores are high
% when they act as a nominal modifier {\sc nn}. This appears due to two reasons:
% \begin{enumerate}
% \item For {\it zebra}, there are frequent erroneous parses such as {\it
%     zebra/{\sc nn} browsing/{\sc root}} instead of {\it zebra/{\sc nsubj} browsing/{\sc
%     root}}. The network does not make this mistake, and treats these
%     occurrences of {\it zebra} according to its importance as {\sc nsubj}.
% \item {\it Water} as a modifier often changes the meaning of its head in
%  a visually salient way: e.g.\ {\it water fall, water balloon, water scene, water skiing}, and
%   thus the network learns that this particular word is important in the modifier position.
% \end{enumerate}

\subsubsection{Sensitivity to linear structure}
\label{subsec:information-struct}

@@

As observed in Section \ref{sec:beyondlexical}, 
adding extra information about the position of words 
explains more of the variance in case of {\sc Visual} and especially
{\sc Textual} and {\sc LM}.
Figure \ref{fig:posrqs} shows the coefficients corresponding to the
position variables in {\sc LR~full}. Since the omission scores 
are measured at the end-of-sentence token, the expectation is that 
for {\sc Textual} and {\sc LM}, as language models,  
the words appearing closer to the end of the sentence would have a
stronger effect on the omission scores. This seems to be confirmed by
the plot as the coefficients for these two networks up until the
\emph{antepenult} are all negative.

For the {\sc Visual} model it is less clear what to
expect: on the one hand due to their chain structure,  
RNNs are better at keeping track of
short-distance rather than long-distance dependencies and thus we can expect 
tokens in positions closer to the end of the sentence to be more important. 
On the other hand in English the information structure a single
sentence is 
expressed via linear ordering: the {\sc topic} of a 
sentence appears sentence-initially, and the {\sc comment} follows. 
In the context of other text types such as dialog or multi-sentence \label{edit:topiccomment}
narrative structure, we would expect {\sc comment} to often be more
important than {\sc topic} as {\sc comment} will often 
contain new information in these cases. In our setting of image captions 
however, sentences are not part of larger discourse, it is sentence
initial material that typically contains the most important 
objects depicted in the image, e.g. {\it {\underline{two zebras}} are grazing in tall grass on a savannah.}
Thus, for the task of predicting features of the visual scene, it would 
be advantageous to detect the topic of the sentence and up-weight its
importance in the final meaning representation. Figure \ref{fig:posrqs}
appears to support this hypothesis and the network does 
learn to pay more attention to  words appearing
sentence-initially. This effect seems to be to some extent mixed with the recency
bias of RNNs as perhaps indicated by the relatively high coefficient of the {\it last}
position for {\sc Visual}.

### 2017-05-30 doc/models.tex

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.25]{imaginet.pdf} 
\caption{Structure of {\sc Imaginet}, taken from Chrupała et al. (2015).}
\label{fig:imaginet}
\end{center}
\end{figure}

@@

Figure~\ref{fig:imaginet} shows the structure of {\sc Imaginet}. As can
be seen from the figure, the model consists of two GRU pathways, 
{\sc Textual} and {\sc Visual}, with a shared word-embedding matrix. 
The inputs to the model are pairs of image descriptions and their 
corresponding images. The {\sc Textrual} pathway predicts the next 
word at each position in the sequence of words in each caption, whereas the 
{\sc Visual} pathway predicts a visual representation of the image that depicts the 
scene described by the caption after the final word is received.

@@

>>>>

We apply our methods to the following models:
\begin{itemize}
\item {\sc Imaginet}: A multi-modal GRU network which consists of two
  pathways, {\sc Visual} and {\sc Textual}, coupled via word
  embeddings.
\item {\sc LM}: A (unimodal) language model consisting of a Gated Recurrent
  Unit (GRU) network.
\item {\sc Sum}: A network with the same objective as the {\sc Visual}
  pathway of {\sc Imaginet}, but which uses sum of word embeddings
  instead of a GRU.
\end{itemize}
The rest of this sections gives more details for these models.

@@

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.2]{imaginet.pdf} 
\caption{Structure of {\sc Imaginet}, adapted from \protect\cite{chrupala2015learning}.}
\label{fig:imaginet}
\end{center}
\end{figure}

Figure~\ref{fig:imaginet} shows the structure of {\sc Imaginet}. As can
be seen from the figure, the model consists of two GRU pathways, 
{\sc Textual} and {\sc Visual}, with a shared word-embedding matrix. 
The inputs to the model are pairs of image descriptions and their 
corresponding images. The {\sc Textual} pathway predicts the next 
word at each position in the sequence of words in each caption, whereas the 
{\sc Visual} pathway predicts a visual representation of the image that depicts the 
scene described by the caption after the final word is received.

@@

\subsection{Unimodal language model}
The model {\sc LM} is a language model analogous to the {\sc Textual}
pathway of {\sc Imaginet} with the difference that its word embeddings
are not shared, and its loss functions is the cross-entropy on word
prediction. Using this model we remove the visual objective as a
factor, as the model does not use the images corresponding to captions
in any way.

\subsection{Sum of word embeddings}
The model {\sc Sum} is a stripped down version of the {\sc Visual}
pathway, which does not share word embeddings, only uses the cosine
loss function, and replaces the GRU network with a summation over word
embeddings. It this removes the effect of word order from
consideration. We use this model as a baseline in the sections which
focus on language structure.
